---
title: "Class 10 p2"
author: "Ruofan Kang"
date: "2023-11-13"
output: pdf_document
---

---
title: "class 10 : Comparative analysis of structure"
author: "Ruofan Kang"
format: pdf
editor: visual
---

We need some packages for todays class. These include 'bio3d' and 'msa'. The 'mea'package is from BioConductor. These packages focus on genomics type work and are managed by the 'BiocManager'package.

Install 'install.packages("BiocManager")' and then 'BiocManager::install("msa")' all entered in the R "brain" console.

```{r}
library(bio3d)

aa <-get.seq("1ake_A")
aa
```

Now I can search the PDB databse for related sequences:

```{r}
b <-blast.pdb(aa)
```

```{r}
hits <- plot(b)
```

```{r}
attributes(b)
head(b$hit.tbl)
```

There are the related structures in the PDB database that we found via a BLAST search...

```{r}
hits$pdb.id
```

Side-note: Lets annotate these structures (in other words find out what thet are, what soeies they are from, stuff about the experiment they were solved in etc.)

For this we can use the 'pdb.annotate()'

```{r}
anno <- pdb.annotate(hits$pdb.id)
```

```{r}
#attributes(anno)
head(anno)
```

Now we can download all these structures for further analysis with the'get.pdb()' function.

```{r}
# Download releated PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

Now we have all these related structures we can Align and Supperpose...

```{r}
# Align releated PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
pdbs
```

## Principal Cpmponent Alalysis

```{r}
# Perform PCA
pc.xray <- pca(pdbs)
plot(pc.xray)
```

```{r}
# Visualize first principal component
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```
