---
title: "class 09"
author: "Ruofan Kang (A17236920)"
format: pdf
editor: visual
---


Today is Halloween and we will apply lots of the analysis methods and R graphics approacches to find out all about typical Halloween candy. 

```{r}
candy_file <- "candy-data.csv"
candy <- read.csv(candy_file, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```


I can convert the 1 and 0 values to be TRUE or FALSE and use that to extract the type of candy I want. For example the chocolate candy...

```{r}
candy[ as.logical(candy$chocolate), ]
```

2.What is your favorate candy?

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy ["Hershey's Milk Chocolate",]$winpercent
```


> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy ["Kit Kat",]$winpercent
```


> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy ["Tootsie Roll Snack Bars",]$winpercent
```


```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Winpercent that looks different scale to the majority of the other columns in the dataset. The most of the variables have values ranging from 0 to 1. However, the variable winpercent mean is 50.32 and other values range from 14 to 84. 


> Q7. What do you think a zero and one represent for the candy$chocolate column?

```{r}
candy$chocolate
```
I think 
Zero means the absence of chocolate in the candy.
One means the presence of chocolate in the candy.


> Q8.Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)

ggplot(candy)+
  aes(winpercent) +
  geom_histogram( binwidth=10 )
```

> Q9. Is the distribution of winpercent values symmetrical?

It is not exactly symmetrical distribution. It looks like tilted to the left. 



> Q10. Is the center of the distribution above or below 50%?

The center of the distribution above 50%. 


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choc.inds <- as.logical(candy$chocolate)
choc.win <- candy [choc.inds,]$winpercent
choc.win
```

```{r}
# Do the same for fruity
fruity.inds <- as.logical(candy$fruity)
fruity.win <- candy [fruity.inds,]$winpercent
fruity.win
```

```{r}
mean(choc.win)
mean(fruity.win)
```


> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruity.win)
```

> Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)
candy %>% arrange(winpercent) %>% head(5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>% arrange(desc(winpercent)) %>% head(5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy)+
  aes(winpercent, rownames(candy)) +
  geom_col(fill="blue")
  
```


> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

##Define some useful colors

```{r}
mycols <- rep("blue", nrow(candy))
#mycols[2:5] <- "red"
mycols [as.logical(candy$fruity)] <- "red"
mycols
```

```{r}
#1 fig-height:10
#1 fig-width:5
ggplot(candy)+
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill=mycols)

```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

Sixlet is the worest ranked chocolate candy.


> Q18. What is the best ranked fruity candy?

Starburst is the best ranked fruity candy. 


```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 15)
```
> Q19.Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures is the highest ranked in terms of winpercent for the least money.


> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```


```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```



```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```


```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
---
