---
title: "Class 06 R functions"
author: "Ruofan Kang (PID:17236920)"
date: "2023-10-19"
output:
  pdf_document
---

# All about functions in R

This week we are introducing **R functions** and how to write our own R functions.
Functions are the way we get stuff in R. We call a function to read data, compute stuff, plot stuff,etc.
R makes writting function accessable but we should always start by tring to get a working snippet of code first before write our function.

## Todays lab
Questions to answer:

>- Q1.Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an NA value) this can be used as a score to be potentially dropped. Your final function should be adquately explained with code comments and be able to work on an example class gradebook such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]

```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)

student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)

student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```


Follow the guidelines from class

- Write a working snipet of code that solves a simple problem

```{r}
# Straight forward mean()
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)

mean(student1)
```
But...We need to drop the lowest socre. First we need identify the lowest score. 

```{r}
# Which element of the vector is the lowest?
which.min(student1)
```
what I want is to now drop (i.e. exclude)this lowest score from my mean() calculation.
```{r}
# This will return everything but the eights
# element of the vector
student1[-8]
```

Now we can use the answer from which.min() to return all other elements of the vector

```{r}
# This is our first working snippet
mean(student1[-which.min(student1)])
```

What about the other example students? Will this work for them? 

We could try using na.rm=TRUE for mean but this is pants!Not a good approach i.e. unfair. 
```{r}
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
mean(student2, na.rm =TRUE)


student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
mean(student3, na.rm =TRUE)

```

Another apprach is to mask (i.e. replace) all NA values with zero

First we need to find the NA elements of the vector.
How do we find the NA elements?
```{r}
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
x<- student2

is.na(x)
which( is.na(x))
```
Now we have identified the NA elements we want to "mask" them. Replace them with zero? 

```{r}
#This does not quite get us there
mean(x[-which(is.na(x))])
```

Instead we will make the NA elements zero

```{r}
#Cool, this is useful!
x[is.na(x)] <-0
x
mean(x)
```

Recall we should drop the lowest score now...

```{r}
x[is.na(x)]<- 0
mean(x[-which.min(x)])
```

Now we are essentialy there with our working snippet!
```{r}
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
x<- student3
x[is.na(x)]<- 0
mean(x[-which.min(x)])
```

## Now we make our function

Take the snippet and trun into a function
Every function has 3 part 

-A name, in our case 'grade()'
-Input arguments, a vector of student scores
-The body i.e. our working snippet of code

Using R Studio I will select 'Code > Extract Function'
```{r}
grade <- function(x) {
  x[is.na(x)]<- 0
  mean( x[-which.min(x)])
}
```

```{r}
grade(student1)
grade(student2)
grade(student3)
```

This looks great! We now need to add comments to explain this to our future selves and other who want to use this function.

```{r}
#' Calculate the average score for a vector of
#' student scores dropping the lowest score.
#' Missing values will be treated as zero.
#' 
#' @param x A numeric vector of homework scores
#' 
#' @return Average score
#' @export
#'
#' @examples
#' student<- c(100,NA,90,97)
#' grade(student)
#'
```

```{r}
grade <- function(x) {
  #mask NA with zero
  #Treat missing values as zero
  x[is.na(x)]<- 0
  #Exclude lowest score from mean
  mean( x[-which.min(x)])
}
```


Now finally we can use our function on our "real" whole class data from this CSV format file:"https://tinyurl.com/gradeinput" 

```{r}
url <-"https://tinyurl.com/gradeinput"
gradebook <- read.csv(url, row.names = 1)
```

```{r}
apply(gradebook,1,grade)
```



>Q2. Using your grade() function and the supplied gradebook, Who is the top scoring student overall in the gradebook? [3pts]

To answer this we run the apply()function and save the rusults. 


```{r}
results <- apply(gradebook,1,grade)
sort(results, decreasing = TRUE)
```

```{r}
which.max(results)
```

>Q3.From your analysis of the gradebook, which homework was toughest on students (i.e. obtained the lowest scores overall? [2pts]

```{r}
gradebook
```


```{r}
ave.scores <- apply(gradebook, 2, mean, na.rm=TRUE)
ave.scores
which.min(ave.scores)
```

```{r}
med.scores <- ave.scores <- apply(gradebook, 2, median, na.rm=TRUE)
med.scores
which.min(med.scores)
```
```{r}
boxplot(gradebook)
```




>Q4.Optional Extension: From your analysis of the gradebook, which homework was most predictive of overall score (i.e. highest correlation with average grade score)? [1pt]

Are the final results (i.e.average score for each student correlated with the results(i.e. scores) for individual homeworks-the gradebook columns)

```{r}
masked.gradebook <- gradebook
masked.gradebook[is.na(masked.gradebook)] <- 0
masked.gradebook
```

And look at corelation

```{r}
cor(results,masked.gradebook$hw5)

```

```{r}
apply(masked.gradebook,2, cor, x=results)
```






>Q5. Make sure you save your Quarto document and can click the “Render” (or Rmark- down”Knit”) button to generate a PDF foramt report without errors. Finally, submit your PDF to gradescope. [1pt]

Knit the document to make a PDF





