---
title: "class 10"
author: "Ruofan Kang (A17236920)"
format: pdf
editor: visual
---

First let's see what is in the PDN database-the mainrepository of protein structures.

Downloaded composition stats from: https://www.rcsb.org/stats/summary

For context: Release 2023_04 of 13-Sep-2023 of UniProtKB/TrEMBL contains 251600768 sequence entries.

http://tinyurl.com/statspdb

```{r}
stats <- read.csv("PDBstats.csv", row.names = 1)
stats
```

There is a problem here due to the commas in the numbers. this cause R to treat them as characters.

```{r}
x<- stats$X.ray
x
```

```{r}
as.numeric( gsub(",", "", x) )
```

```{r}
rm.comma <- function(x) {
  as.numeric( gsub(",", "", x) )
}
```

```{r}
rm.comma(stats$X.ray)
```

I can use 'apply()' to fix the whole table...

```{r}
pdbstats <- apply(stats, 2, rm.comma)
rownames(pdbstats) <-rownames(stats)
head(pdbstats)
```

```{r}
totals <-apply(pdbstats, 2, sum)
totals
```

> Q1: What percentage of structures in the PDB are solved by X-Ray and Electron Microscopy.

```{r}
round(totals/totals["Total"] *100,2)
```

> Q2: What proportion of structures in the PDB are protein?

```{r}
round(pdbstats[1, "Total"]/ sum(pdbstats [, "Total"]) * 100, 2)
```

> Q3: Type HIV in the PDB website search box on the home page and determine how many HIV-1 protease structures are in the current PDB?

SKIPED for time

Protein structures in PDB as a fraction of Uniprot sequences. See:https://www.uniprot.org/help/release-statistics

```{r}
round( (pdbstats[1, "Total"]/ 251600768)*100,2)
```

> Q4: Water molecules normally have 3 atoms. Why do we see just one atom per water molecule in this structure?

This is a 2 Angstrom structure and hydrogen is not visible at this resolution in X-ray crystallography. The positions of the hydrogen atoms are usually not determined, their electron density is weak and harder to discern.

> Q5: There is a critical "conserved" water molecule in the binding site. Can you identify this water molecule? What residue number does this water molecule have

Water HDH 308

> Q6: Generate and save a figure clearly showing the two distinct chains of HIV-protease along with the ligand. You might also consider showing the catalytic residues ASP 25 in each chain and the critical water (we recommend "Ball & Stick" for these side-chains). Add this figure to your Quarto document.

Here is a lovely figure of HIP-Pr with the catalytic ASP residues,the MK1 compound and the all important water 308.

```{r}
plot(magick::image_read("1HSG-2.png"))
```

```{r}
plot(magick::image_read("1HSG-3.png"))
```

## The bio3d package for structural bioinformatics

```{r}
library(bio3d)
pdb <- read.pdb("1hsg")
pdb
```

```{r}
print(pdb)
```

```{r}
read.pdb(file="1hsg")
```

> Q7: How many amino acid residues are there in this pdb object?

There are 198 amino acid residues.

> Q8: Name one of the two non-protein residues?

HOH

> Q9: How many protein chains are in this structure?

2 protein chains are in this structure.

```{r}
attributes(pdb)
```

```{r}
head(pdb$atom)
```

#Predicting functional motions of a single structure

Let's finish today with a bioinformatics calculation to predict the functional motions of a PDB structure.

```{r}
adk <- read.pdb("6s36")
m <- nma(adk)
```

```{r}
plot(m)
mktrj(m, file="adk_m7.pdb")
```

> Q10. Which of the packages above is found only on BioConductor and not CRAN?

msa package is found only on Bioc\]Conductor not CRAN.

> Q11. Which of the above packages is not found on BioConductor or CRAN?

"Grantlab/bio3D-view" package is not found on BioConductor or CRAN.

> Q12. True or False? Functions from the devtools package can be used to install packages from GitHub and BitBucket?

True.

```{r}
library(bio3d)
aa <- get.seq("1ake_A")
```

```{r}
aa
```

> Q13. How many amino acids are in this sequence, i.e. how long is this sequence?

This sequence is 214 amino acids long.
